---
title: "Social Network Analysis: 2022"
subtitle: "Home Assignment 2"
author: Kirill Guliev, MA Comparative Social Research 
date: "06 Feb 2022"
output:
  word_document:
    css: "my-style.css"
toc: true
---


```{r setup, include = FALSE, echo = FALSE}

knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE,
                      fig.width = 7,
                      fig.height = 4.5)

path <- "C:/Users/Кирилл/Desktop/CSR2023/SNA/social-network-analysis-2022"

```

# Loading libs needed to answer assignment questions

```{r libraries, echo = TRUE}

library(sna)
library(network)
library(igraph)
library(RColorBrewer)

```

# Answers on Assignment questions

## Question № 1: Why did we use option FALSE for command “directed” above (_below_), when creating a network?

```{r question 1, echo = TRUE}


# path is specified in the very first chunk which is hidden

load(paste0(path, "/data/madmen.Rdata"))

mad.net <- as.network(mad.matrix, directed = FALSE)

network::set.vertex.attribute(mad.net, attrname = "female", value = mad.att[, 2])


```

## Answer: 

We define the argument `directed` to `FALSE` because the relations between MadMen TV-show are not directed but reciprocal. 

## Question 2: Plot madmen network using node shapes for different genders

## Answer:

```{r males-females, echo = TRUE}

# How many males and females in our data?

table(
  mad.att[, 2]
)

# 15 males and 30 females

```

We have 15 males and 30 females in a network.

```{r madmen plotting, echo = TRUE}

gender <- mad.att[, 2]

sides <- ifelse(gender == 0, 7, 3)

par(mar = c(0, 0, 0, 0))
plot(
  mad.net,
  vertex.sides = sides,
  vertex.cex = 1.5
)

```

## Question 3: Draw one of Florentine networks in different layouts. What do you observe?

## Answer:

For Kamada & Kawai and Fruchetrman & Reingold layouts we observe that the length of edges is approximately equal for the whole network. Moreover, we see that edges in the network do not intersect (as these parameter is minimized by the algorithm). In the `circle` layout we see that nodes are situated in the plot in the circle pattern, edges are of different lengths, and there are a lot of intersections that make the plot difficult to read and analyze. 

```{r florentine networks, echo = TRUE}

load(paste0(path, "/data/flo.Rdata"))

flo.marriage <- as.network(as.matrix(flo.marriage), directed = FALSE)

flo.biz <- as.network(as.matrix(flo.biz), directed = FALSE)

# Add attributes
network::set.vertex.attribute(flo.marriage, "wealth", flo.att[, 2])
network::set.vertex.attribute(flo.biz, "wealth", flo.att[, 2])

# Plots

par(mar = c(0, 0, 0, 0))

plot(
  flo.marriage,
  vertex.col = network::get.vertex.attribute(flo.biz, 'wealth') / 25 + .4,
  vertex.cex = 1.5,
  display.labels = TRUE,
  mode = "fruchtermanreingold"
)

plot(
  flo.marriage,
  vertex.col = network::get.vertex.attribute(flo.biz, 'wealth') / 25 + .4,
  vertex.cex = 1.5,
  display.labels = TRUE,
  mode = "kamadakawai"
)

plot(
  flo.marriage,
  vertex.col = network::get.vertex.attribute(flo.biz, 'wealth') / 25 + .4,
  vertex.cex = 1.5,
  display.labels = TRUE,
  mode = "circle"
)

```

## Question 4: Generate a set of constant coordinates, and draw both marriage and business Florentine networks in the same coordinates.

## Answer:

```{r constant coordinates, echo = TRUE}

nodes_n <- network.size(flo.marriage)

x <- sample((0:(nodes_n - 1)) / nodes_n)

y <- sample(x, nodes_n)

coords <- cbind(x, y)

# for Kamada & Kawai algorithm

coord_kk <- network.layout.kamadakawai(flo.marriage, layout.par = list(seed.coord = coords))
coord_kk_biz <- network.layout.kamadakawai(flo.biz, layout.par = list(seed.coord = coords))

par(mar = c(0.5, 0.5, 0.5, 0.5), mfrow = c(1, 2))

plot(
  flo.marriage,
  displaylabels = TRUE,
  label.cex = .4,
  label.pos = 1,
  coord = coord_kk
)

plot(
  flo.biz,
  displaylabels = TRUE,
  label.cex = .4,
  label.pos = 1,
  coord = coord_kk_biz
)

```

# Answers on Homework Assignment

## 1.  Examine the drug network we’ve drawn in the seminar 3. Does anything look odd to you?

The odd thing about the drug network is that there are a lot of missed connections - in other words, a lot of nodes that do not have edges with others. So, it happens that these observations were reported as drug users but have no ties with others.  

## 2. For the drug network, draw the network with nodes colored by race. What do you conclude?

We see two clusters - for the 2nd and 3d ethnicity codes. These drug users are connected not only by their habit of using drugs but also by their ethnic identity. 

```{r second question, echo = TRUE}


drugpaj <- read.paj(paste0(path, "/data/drugnet2.paj"))
drug <- drugpaj$networks[[1]] # Extract network

# Extract attributes:
ethnicity <- drugpaj$partitions[[2]]

table(ethnicity)

#Set colors by gender, including gray for undecided:
colors <- dplyr::case_when(
  ethnicity == 1 ~ "coral3",
  ethnicity == 2 ~ "darkolivegreen1",
  ethnicity == 3 ~ "darkslategray1",
  ethnicity == 4 ~ "hotpink"
)

par(mar = c(0, 0, 0, 0)) # And the plot itself:
plot(
  drug,
  vertex.col = colors,
  vertex.cex = 1.5
)

```

## 3. Please examine the options in the "network.layout" command and perform the following on the madmen data

### 3.1 Create the madmen.net with labels

```{r madmen with lables, echo = TRUE}
par(mar = c(0, 0, 0, 0))
plot(
  mad.net,
  displaylabels = TRUE
)

```

### 3.2 Experiment with options by adding attributes, changing vertex or edge colors, finding the best position for labels

```{r experiment, echo = TRUE}

set.seed(2022)
par(mar = c(0, 0, 0, 0))
plot(
  mad.net,
  displaylabels = TRUE,
  label.pos = 0,
  label.cex = .4,
  vertex.cex = 1.5,
  vertex.sides = sides,
  vertex.col = sample(c("red", "blue", "yellow",
                        "lightgreen", "mediumpurple2",
                        "magenta2", "palevioletred3",
                        "turquoise1", "steelblue"), 4)
)

```

## 4. Think about the manipulations we are performing on the trade network. With respect to dichotomizing, please answer the following

### 4.1 How would you justify any of these choices?

- In case of the first option, I would justify it if I had to define whether the country has or not any trade relations with other world arena actors. Because 1 represents that there is a relationship (because the value is greater than 0) and 0 represents the absence of such relationship. 

- In case of the second option, I would justify it if I had to answer the question about relative consistency of relations between actors. It means that if actor has more than 2 (included) trade-related conncetions, we should consider it consistent. 

- In case of the third option, I would justify this dichotomy if I had to answer the question to display actors who has the strongest connection and relationships in trading sphere. 

### 4.2 What are the empirical emplication of these choices?

The empirical implication is that we receive absolutely different graphs and, therefore, get different insights. With the first option we draw and overall picture of trade relations; with the second - we depict some consistent and somehow moderate relationships; and with the final option we draw a picture of major allies in trade relations. 

### 4.3 What differences do you observe between the graphs where the cutpoint is any tie, at least two ties, and all ties present?

The major difference is that in the last graph where the condition of dichotomy was extreme (all ties are present) the information can be read without difficulties - it is clear and easy to get what country has all trade-related connection with other countries; while in the first two graphs it is tricky to distinguish those ties.  

### 4.4 What information can you gather from these observed differences to help you expand on your earlier theoretical justification of ties? Alternatively, does your theoretical justification seem reasonable in light of new information obtained from these graphs?

Dichotomy must be applied only after deep theoretical consideration which should be agreed-upon with research goals, question, and problem.  

## 5. select one of the individual trade networks (manufacture, food, crude, etc.) and show me everything you’ve learned in this class so far 

### 5.1 Create an appropriate graph with all possible options

```{r network creation, echo = TRUE}

load(paste0(path, "/data/trade.Rdata"))

minerals_mat <- as.matrix(minerals)

energy <- ifelse(trade.att[, 4] > median(trade.att[, 4], na.rm = TRUE), 3, 4)

# 3 - triangle
# 4 - square

#head(minerals_mat, 4)

minerals_net <- as.network(minerals_mat, directed = TRUE)

network::set.vertex.attribute(minerals_net, attrname = "ENERGY", value = trade.att[, 4])

set.seed(2222)
par(mar = c(0, 0, 0, 0))
plot(
  minerals_net,
  displaylabels = TRUE,
  label.cex = .5,
  label.pos = 1,
  vertex.cex = 1.5,
  vertex.col = network::get.vertex.attribute(minerals_net, "ENERGY") / 25 + .4,
  vertex.sides = energy,
  mode = "kamadakawai"
)

```

### 5.2 Tell me what inferences you can make about your selected network based on the information you’ve obtained. Supplement your arguments with logic and theory.

In the graph above, the shape of nodes (triangles and squares) represents the volume of energy import/export. If it is triangle, it means that the country has the volume of energy imported/exported greater than the median. If it is square, it means that the volume is lower than the median. We see that "triangle" countries are mostly those which can be considered as *developed* ones; and "square" countries are those which are developing (or third-world countries). It is logically correct that developed countries import and/or export energy resources to and from developing countries. Also, this feature is connected with the data on minerals (including oil) export/import upon which this network was created. Network's density is equal to 0.24 (was received from `summary(minerals_net)` call).