---
title: "Social Network Analysis: 2022"
subtitle: "Home Assignment 1"
author: Kirill Guliev, MA Comparative Social Research 
date: "30 Jan 2022"
output: word_document
toc: true
---

```{r, echo = FALSE}

knitr::opts_chunk$set(message = F, warning = FALSE)

```

# Loading all vital SNA libaries into the environment

```{r}

library(tidyverse)
library(network)
library(igraph)
library(RColorBrewer)

```

# Code from the seminar 

```{r, fig.width=6, fig.height=6}
drugpaj <- read.paj('C:/Users/Кирилл/Desktop/CSR2023/SNA/social-network-analysis-2022/data/drugnet2.paj')

names(drugpaj) # objects in drugpaj

names(drugpaj$networks) 

names(drugpaj$partitions)

drug <- drugpaj$networks[[1]] # extract network

class(drug)

plot(drug)

```

## Question 1: What looks strange about this network and why?

```{r}

#summary(drug)

network.size(drug) # how many nodes?
network.edgecount(drug) # how many edges?
network.dyadcount(drug) # how many dyads?

```
### Answer:

There are elements in this network that are not connected with each other. We observe, that the number of these elements (_nodes_, `r network.size(drug)`) is greater than the number of edges in this network (`r network.edgecount(drug)`). 

## Question 2: What do the numbers above represent? (Dyads, nodes, edges)

```{r, eval = FALSE}

network.size(drug) # how many nodes?
network.edgecount(drug) # how many edges?
network.dyadcount(drug) # how many dyads?

```

### Answer:

* The first number - `r network.size(drug)` - represents the number of units of analysis. Say, these are people, drug users, who are under researcher's investigation.

* The second number - `r network.edgecount(drug)` - represents the number of edges; or, in other words, the number of relationships between nodes. 

* The third number - `r network.dyadcount(drug)` - represents the number of dyads, which is basically a relationship between two nodes. 

## Question 3:  Plot the biz network with node attributes that you’ve set above

```{r, fig.width=7, fig.height=3.5}
# Seminar code

load('C:/Users/Кирилл/Desktop/CSR2023/SNA/social-network-analysis-2022/data/flo.Rdata')

flo.marriage <- as.network(as.matrix(flo.marriage), directed = FALSE)
flo.biz <- as.network(as.matrix(flo.biz), directed = FALSE)
# Add attributes
network::set.vertex.attribute(flo.marriage, 'wealth', flo.att[, 2])
network::set.vertex.attribute(flo.biz, 'wealth', flo.att[, 2])
# Simple plots:
par(mar = c(0, 0, 0, 0))
plot(flo.marriage)
plot(flo.biz)

```


### Answer:

```{r, fig.width=7, fig.height=3.5}

par(mar = c(0, 0, 0, 0))
plot(
  flo.biz,
  vertex.cex = (network::get.vertex.attribute(flo.biz, 'wealth') / 25 + .4),
  displaylabels = TRUE,
  label.cex = .5,
  label.pos = 0,
  coord = plot(flo.biz)
)

```


## Question 4: Using the code already shown, plot both of the new networks. Add attributes if you wish


```{r}
# Read vertex labels with scan()
flo.names <-
  scan(
    'C:/Users/Кирилл/Desktop/CSR2023/SNA/social-network-analysis-2022/data/padgett.dat',
    what = 'character',
    skip = 4,
    nlines = 16
  )
# Read data with read.table()
flos <-
  read.table(
    'C:/Users/Кирилл/Desktop/CSR2023/SNA/social-network-analysis-2022/data/padgett.dat',
    skip = 41,
    col.names = flo.names
  )
# Read node attributes:
flo.att <-
  read.table(
    'C:/Users/Кирилл/Desktop/CSR2023/SNA/social-network-analysis-2022/data/padgw.dat',
    col.names = c('WEALTH', 'NUM.PRIORS', 'NUM.TIES'),
    skip = 25
  )


# Separate adjacency matrices
flo.marriage <-
  flos[1:16, ]  # subset of the first 16 colums is the marriage network
dim(flo.marriage)
row.names(flo.marriage) <- flo.names # name
flo.biz <-
  flos[17:32, ] # subset of the second 16 is the business network.
row.names(flo.biz) <- flo.names # name
dim(flo.biz)
# Check the data by listing a couple of rows and colums from each network.
flo.marriage[1:2, 1:2]
flo.marriage[15:16, 15:16]
flo.biz[1:2, 1:2]
flo.biz[15:16, 15:16]

flo.marriage <- as.network(as.matrix(flo.marriage), directed = FALSE)
flo.biz <- as.network(as.matrix(flo.biz), directed = FALSE)

## add attributes
network::set.vertex.attribute(flo.marriage, 'wealth', flo.att[, 2])
network::set.vertex.attribute(flo.biz, 'wealth', flo.att[, 2])
```


### Answer:

```{r, fig.width=7, fig.height=3.5}

par(mar = c(0, 0, 0, 0))
plot(
  flo.biz,
  vertex.cex = (network::get.vertex.attribute(flo.biz, 'wealth') / 25 + .4),
  displaylabels = TRUE,
  label.cex = .5,
  label.pos = 0,
  coord = plot(flo.biz)
)
plot(
  flo.marriage,
  vertex.cex = (network::get.vertex.attribute(flo.marriage, 'wealth') / 25 +
                  .4),
  displaylabels = TRUE,
  label.cex = .5,
  label.pos = 0,
  coord = plot(flo.marriage)
)

```

## Question 5: For the network "drug" that we created and loaded with attributes, create several different network plots, adding gender and ethnicity to the graph as node attributes. 

### Answer:

```{r, fig.width=7, fig.height=3.5}

names(drugpaj$partitions) # attributes included
gender <- drugpaj$partitions[[1]] # extract gender
#gender #check the values assigned to each gender
table(gender) # how many women/men
# It is actually better to recode gender into 1 and 0 as opposed to 2 and 1;
# setting 0/1 variable as an attribute is cleaner that way.
female <- ifelse(gender == 2, 1, # recode
                 ifelse(gender == 1, 0, NA))
## set attributes
drug <-
  network::set.vertex.attribute(drug, 'female', value = c(female))
ethnicity <- drugpaj$partitions[[2]] # extract ethnicity
table(ethnicity) # how is it coded?
drug <-
  network::set.vertex.attribute(drug, 'ethnicity', value = c(ethnicity))


par(mar = c(0, 0, 0, 0))
plot(
  main = "Drug users by ethnicity",
  drug,
  vertex.col = (network::get.vertex.attribute(drug, 'ethnicity')),
  displaylabels = TRUE,
  label.cex = .5,
  label.pos = 0,
  coord = plot(drug)
)
plot(
  main = "Drug users by gender",
  drug,
  vertex.col = c("red", "blue"),
  vertex.cex = (network::get.vertex.attribute(drug, 'female')),
  displaylabels = TRUE,
  label.cex = .5,
  label.pos = 0,
  coord = plot(drug)
)
```